---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tractr <img src="man/figures/tractr-logo.svg" align="right" height="139"/>

<!-- badges: start -->
<!--insert CRAN status badge -->
<!--insert R-CMD-check badge -->
[![Codecov test coverage](https://codecov.io/gh/richford/tractr/branch/main/graph/badge.svg)](https://app.codecov.io/gh/richford/tractr?branch=main)
<!-- badges: end -->

## Overview

tractr contains functions to analyze diffusion MRI (dMRI) tract profiles using generalized additive models. It operates on outputs from the AFQ software ecosystem and tests for differences between two phenotypic groups along the length of a specified white matter tract.

## Installation

You can install the development version of tractr from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("richford/tractr")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(tractr)
## basic example code
```
